# Exercise 5: Dimensionality reduction techniques

*November 29, 2019*  
*This week I performed the dimensionality reduction techniques on "human" data.*  

### {.tabset}

#### **Playing with the dataset**  

**Loading** the "human" data: 
```{r}
library(MASS)
human <- read.csv("./data/human.csv", row.names = 1)
str(human)
summary(human)
```
It is a United Nations dataset for Human Development Index in different countries depending on such parameters as education, labour, life expectancy and others.  
The dataset has **195 observations** (rows) and **19 variables** (columns).  

Let's look at a graph:  
```{r}
pairs(human)
```

We can see scatterplots of the relationships of data’s variables.  
We can see some U-shaped patterns and a lot of linear patterns that show that the variables are probably related. Making examples here would be exuberating, so lets proceed to look at statistics as it is. 

#### **Principal component analysis**  

#### Non-standardized variables
```{r}
library(dplyr)
library(stringr)
human <- mutate(human, GNI = str_replace(human$GNI, pattern=",", replace ="") %>% as.numeric)
pca_human <- prcomp(human)
```

**Variability** captured by PCA:
```{r}
summary(pca_human)
```

Now let's draw **a biplot**:
```{r }
biplot(pca_human, xlab = pca_human$PC1, ylab = pca_human$PC2, main = "Non-standardized variables: GNI and others")
```

It looks pathetic. Lets try to scale the variables now.

#### Standardized variables
```{r}
human_std <- scale(human)
summary(human_std)
pca_human_std <- prcomp(human_std)
summary(pca_human_std)
biplot(pca_human_std, col = c("grey40", "deeppink2"), xlab = pca_human_std$PC1, ylab = pca_human_std$PC2, main = "Standardized variables: mostly ducation and labour")
``` 

Looks much better! (We've also changed the color).  
The scaling normalizes the range of variables, so it looks more neat.  

**Interpretations**  

Based on a scaled bi-plot, here is what we see:  

1. **The angle between features** represents high correlation between them.  
- Expected years of schooling (Edu.Exp), Life expectancy at birth (Life.Exp), Proportion of females and males with at least secondary education (Edu2.FM) and Gross National Income per capita (GNI) are *highly positively correlated*.  
- Percentage of female representatives in parliament (Parli.F) and Proportion of females and males in the labour force (Labo.FM) are also *highly positively correlated*. However, this correlation is smaller than (a) and (c).
- Adolescent birth rate (Ado.Birth) and Maternal mortality ratio (Mat.Mor) are also *highly positively correlated*.  

2. **The angle between the feature and the PC axis** represents correlation between the two. If PC and the feature arrow are "in parallel", it means, they are contributing to the same dimension.
- Percentage of female representatives in parliament (Parli.F) and Proportion of females and males in the labour force (Labo.FM) contribute to the PC2.
- All other features contribute to PC1.  

#### **Tea dataset** 

```{r}
library(FactoMineR)
data("tea")
str(tea)
dim(tea)
```

Now lets select columns to do MCA:  
```{r}
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
tea_time <- dplyr::select(tea, one_of(keep_columns))
```

**Time for MCA!**
```{r}
mca <- MCA(tea_time, graph = FALSE)
summary(mca)
```

**Interpretation**:  

The table contains the Eigenvalues, the Individuals, Categories and Categorical variables.  

Let look at categorical variables: if the value is close to 1, then it means a strong link with the variable and dimension. We can see it here in "where" and "Dim2" and "Dim1". Also in "how" and "Dim1". 

Now lets **draw an MCA variable biplot**:
```{r}
plot(mca, invisible=c("var"))
```

**Interpretation:**  

Variables are shown here on the first two dimensions. We can look at possible variable patterns: the distances give measures of variables similarity: the closer - the better.  

For me, the variable plot makes no sense, so I prefer to **draw an individuals plot**.  

```{r}
plot(mca, invisible=c("ind"))
```

Here, for example, "unpackaged" and "tea shop" are more similar, than "lemon" and "black".  


